---
title: "Lab 6"
author: "Sakib Salim"
output: pdf_document
date: "11:59PM March 24, 2019"
---

Load the Boston Housing data and create the vector $y$ and the design matrix $X$.

```{r}
data(Boston, package = "MASS")
y = Boston$medv
X = rep(1, nrow(Boston))
X = as.matrix(cbind(X, Boston[, 1 : 13]))
```

Find the OLS estimate and OLS predictions without using `lm`.

```{r}
b = solve(t(X) %*% X) %*% t(X) %*% y
yhat = X %*% b
```

Write a function spec'd as follows:

```{r}
#' Orthogonal Projection
#'
#' Projects vector a onto v.
#'
#' @param a   the vector to project
#' @param v   the vector projected onto
#'
#' @returns   a list of two vectors, the orthogonal projection parallel to v named a_parallel, 
#'            and the orthogonal error orthogonal to v called a_perpendicular
orthogonal_projection = function(a, v){
  a_parallel = (v %*% t(v) %*% a) / ( sum(v^2) )
  a_perpendicular = a - a_parallel
  list("a_parallel" = a_parallel, "a_perpendicular" = a_perpendicular)
}

```


Try to project onto the column space of $X$ by projecting $y$ on each vector of $X$ individually and adding up the projections. You can use the function `orthogonal_projection`.

```{r}
sumOrthProj = rep(0, nrow(X))
for (j in 1 : ncol(X)){
  sumOrthProj = sumOrthProj + orthogonal_projection(y, X[, j])$a_parallel
  }
head(sumOrthProj)
```

How much double counting occurred? Measure the magnitude relative to the true LS orthogonal projection.

```{r}
doublecount = sumOrthProj / yhat
head(doublecount)
```

Convert $X$ into $Q$ where $Q$ has the same column space as $X$ but has orthogonal columns. You can use the function `orthogonal_projection`. This is essentially gram-schmidt.

```{r}
Q = matrix(NA, nrow = nrow(X), ncol = ncol(X))
Q[, 1] = X[, 1]
for(j in 2 : ncol(X)){
  Q[ , j] = X[ , j]
  
  for(j0 in 1 : (j - 1)){
    Q[ , j] = Q[ , j] - (orthogonal_projection(X[ , j], Q[ , j0])$a_parallel)
  }
}
pacman::p_load(Matrix)
rankMatrix(Q)[1]
```

Make $Q$'s columns orthonormal.

```{r}
for (j in 1 : ncol(Q)){
  Q[ ,j] = Q[ , j] / sqrt(sum(Q[ ,j]^2)) 
}
```

Verify $Q^T$ is the inverse of $Q$.

```{r}
#TO-DO
head(t(Q) %*% Q)
```


Project $Y$ onto $Q$ and verify it is the same as the OLS fit.

```{r}
head(cbind(Q %*% t(Q) %*% y, yhat))
```


Project $Y$ onto the columns of $Q$ one by one and verify it sums to be the projection onto the whole space.

```{r}
#TO-DO
proj_col_Q = rep(0,ncol(Q))
for (j in 1 : ncol(Q)){
  proj_col_Q = proj_col_Q + orthogonal_projection(y, Q[, j])$a_parallel
}
proj_Q = Q %*% t(Q) %*% y
pacman::p_load(testthat)
expect_equal(proj_col_Q, proj_Q)
```

Verify the OLS fit squared length is the sum of squared lengths of each of the orthogonal projections.

```{r}
#TO-DO
sum_sq_length_col_Q = 0
for (j in 1 : ncol(Q)){
  col_proj = orthogonal_projection(y, Q[, j])$a_parallel
  sum_sq_length_col_Q = sum_sq_length_col_Q + sum(col_proj^2)
}
OLS_sq_length = sum(proj_Q^2)
expect_equal(sum_sq_length_col_Q, OLS_sq_length)
```

Rewrite the "The monotonicity of SSR" demo from the lec06 notes but instead do it for RMSE. Comment every line in detail. Write about what the plots means.

```{r}
#TO-DO
n = 100
y = rnorm(n)
Rmses = array(NA, n) #array to store RMSE values

#Create data matrix
X = matrix(NA, nrow = n, ncol = 0)

#for every new p, add a random column
for (p_plus_one in 1 : n){          #loop through eventual columns of X
  X = cbind(X, rnorm(n))            #add a random column
  Rmses[p_plus_one] = summary(lm(y ~ X))$sigma  #find RMSE
}
pacman::p_load(ggplot2)
base = ggplot(data.frame(p_plus_one = 1 : n, RMSE = Rmses))
base + geom_line(aes(x = p_plus_one, y = RMSE))
#The RMSE value decreases as we add columns to our data.
#Eventually R^2 approaches 1 which in turn makes RMSE = 0.
```

Rewrite the "Overfitting" demo from the lec06 notes. Comment every line in detail. Write about what the plots means.

```{r}
#TO-DO
bbeta = c(1, 2, 3, 4)   #the real betas

#build training data
n = 100
X = cbind(1, rnorm(n), rnorm(n), rnorm(n)) #Intercept plus 3 random columns
y = X %*% bbeta + rnorm(n, 0, 0.3) #ma

#build test data
n_star = 100
X_star = cbind(1, rnorm(n), rnorm(n), rnorm(n_star))
y_star = X_star %*% bbeta + rnorm(n, 0, 0.3)

#store the betas each time you model on 
#a design matrix with p+1 columns
all_betas = matrix(NA, n, n)
all_betas[4, 1 : 4] = coef(lm(y ~ 0 + X)) #fourth row of beta matrix are the 
#beta values from when we had 4 columns in X
in_sample_rmse_by_p = array(NA, n)        #Store In Sample RMSE
for (j in 5 : n){
  X = cbind(X, rnorm(n))                  #add a random column to X
  lm_mod = lm(y ~ 0 + X)                  #new linear model w/o intercept
  
  all_betas[j, 1 : j] = coef(lm_mod)      #add betas to the beta matrix in the 
  #appropriate column (for j columns add them to the jth column of beta matrix)
  
  y_hat = X %*% all_betas[j, 1 : j]                   #produce a prediction our 
  #new model(using new betas)
  
  in_sample_rmse_by_p[j] = sqrt(sum((y - y_hat)^2))   #calculate RMSE for the 
  #prediction of "each new linear model"
}
plot(1 : n, in_sample_rmse_by_p)
#RMSE decreases as the columns of 
#our design matrix approach n
#not honest, since we are just testing on given data,
#so eventually our model passes through every point in our data

head(all_betas[4 : n, 1 : 4]) #take the first four betas for each of our models
b_error_by_p = rowSums((all_betas[, 1 : 4] - matrix(rep(bbeta, n), nrow = n, byrow = TRUE))^2) 

#compare the model betas to the real betas
plot(1 : n, b_error_by_p)
#our betas degenerate as the number of columns approaches n

#look at out of sample error in the case of only the first four features
oos_rmse_by_p = array(NA, n)    #store oosRMSE
for (j in 4 : n){
  y_hat_star = X_star %*% all_betas[j, 1 : 4]    
  #predict on the data using the betas from "first four columns of Xstar" 
  oos_rmse_by_p[j] = sqrt(sum((y_star - y_hat_star)^2)) #calculate RMSE
}
plot(1 : n, oos_rmse_by_p)

#look at out of sample error 
#in the case of the random features too
oos_rmse_by_p = array(NA, n)
for (j in 5 : n){
  X_star = cbind(X_star, rnorm(n))            #again add a random column
  y_hat_star = X_star %*% all_betas[j, 1 : j] #prediction
  oos_rmse_by_p[j] = sqrt(sum((y_star - y_hat_star)^2)) #oosRMSE for when you add random columns
}
plot(1 : n, oos_rmse_by_p)
#oosRMSE actually tests how well our model 
#works on data outside of the given data
#the predicting power of our model 
#significantly decreases as we increase the columns of X
```
